---
title: "State of the U.S. Credit Market: August 2023"
author: "Scott Monaco"
date: "8/29/2022"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
urlcolor: blue
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(dev='CairoPDF', message=FALSE, warning=FALSE, echo = FALSE, fig.width = 7, fig.height = 4)
if(!require("pacman")) install.packages("pacman")
pacman::p_load(data.table, ggplot2, dplyr, tidyr, quantmod, ggrepel, 
               reshape2, readxl, scales, gghighlight, cairo_pdf)
```

```{r echo=FALSE, warning=FALSE}
# load fonts
source('../theme/wharton_theme.R')
```

```{r echo=FALSE, warning=FALSE}
# load function
dtfmt <- function(x){
  parts = strsplit(x, ':')
  # print(parts)
  # print(paste0("20", parts[[1]][1], ' ', parts[[1]][2]))
  return(as.yearqtr(paste0("20", parts[[1]][1], ' ', parts[[1]][2])))
}
```

\pagebreak

# Summary of the U.S. Credit Market 

One of the biggest debates in the market today is over what exactly is going to happen with consumer credit. Delinquencies on various kinds of debt are rising, but the rate at which people are falling into outright distress remains low. How long can that continue?

Until very recently, the rate of late payments had trended below historical norms across virtually all lending types and consumer groups. A frequent explanation is that Americans built up a cushion of extra cash savings during the pandemic and still haven’t spent it down.  

```{r, results = TRUE, echo = FALSE, warning=FALSE}
setwd("~/python_scripts/git/market_research_jan_2022/2023-08-update")
dtapath <- "../credit_data/HHD_C_Report_2023Q2.xlsx"

credbal <- read_excel(path = dtapath, sheet = "Page 3 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`...1`)
credbal$timeper2 <- as.yearqtr(sapply(credbal$timeper, dtfmt))
credmelt <- credbal %>% select(-Total, -timeper) %>% melt(id.vars = c('timeper2'))

ggplot(credmelt, aes(x=timeper2, y=value, fill=variable)) +
  geom_col() +
  labs(
    title = "Total Consumer Debt Balance & Composition"
    , x = "Year : Quarter"
    , y = "Outstanding Balance ($ trillion)"
    , fill = "Debt Type"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=dollar) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values = c(wharton_palette[c(4,2)], wharton_palette[1], wharton_palette[c(5,6,7)]))
# addlogo()


```

CC balances way up
```{r, results = TRUE, echo = FALSE, warning=FALSE}
#credit limit and balance
cclimit <- read_excel(path = dtapath, sheet = "Page 10 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`...1`) 
cclimit$timeper <- as.yearqtr(sapply(cclimit$timeper, dtfmt))
#delete every even row
toDelete <- seq(0, length(cclimit$timeper),2)
cclimit <- cclimit[-toDelete, ]

#just cc balance (geom line)
ggplot(cclimit, aes(x=timeper, y=`Credit Card Balance`, group=1)) +
  geom_line(color=wharton_palette[4], size=1) +
  geom_area(fill=wharton_palette[4], alpha=0.5) +
  labs(
    title = "Credit Card Balances, Quarterly"
    , x = "Year : Quarter"
    , y = "Balance ($T)"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=dollar) +
  # scale_x_date(date_labels = '%Y', date_breaks = '1 year') +
  scale_x_yearqtr(format = '%Y-Q%q') +
  wharton_theme
# addlogo()

```

Business applications
```{r, results = TRUE, echo = FALSE, warning=FALSE}
# new business applications
getSymbols.FRED("BABATOTALSAUS", env = .GlobalEnv) 

new_biz_apps <- as.data.frame(BABATOTALSAUS)
new_biz_apps$dt <- as.Date(row.names(new_biz_apps))

new_biz_apps$yrqtr <- as.yearqtr(new_biz_apps$dt)

ggplot(new_biz_apps%>%filter(dt>='2005-01-01'), aes(x=yrqtr, y=BABATOTALSAUS, group=1)) +
  geom_line(color=wharton_palette[4], size=2) +
  # geom_point(data = ueclaims%>%filter(dt >= '2020-04-01'), color=wharton_palette[7]) +
  # geom_text_repel(data = ueclaims%>%filter(dt >= '2020-04-01'),
  #                 aes(label=paste0(strftime(dt, '%b %d'), ": ", comma(CCSA))),
  #                 size = 3.5, family='Source Sans Pro') +
  labs(
    title = "New Business Applications (total for all NAICS in U.S.)"
    , x = "Year"
    , y = "Number of New Business Applications"
    , caption = "Data:U.S. Census Bureau [BABATOTALSAUS]"
  ) +
  scale_y_continuous(labels=comma) +
  # scale_x_date(date_labels = '%Y-%m', date_breaks = '1 year') +
  scale_x_yearqtr(format = '%Y-Q%q') +
  wharton_theme 
# addlogo()

```

## Rising Delinquencies

Yet the frequency with which people are becoming late with payments on their debts for some kinds of loans is returning not just to prepandemic levels, but even moving beyond them. The percentage of credit-card and auto-loan balances transitioning into delinquency—that is, going from current to becoming 30-days-plus late—is happening at a pace faster than that of 2019, according to the Federal Reserve Bank of New York’s recently released second-quarter Quarterly Report on Household Debt and Credit, which is based on a nationally representative sample of anonymized Equifax credit data.


```{r, results = TRUE, echo = FALSE, warning=FALSE}
#Transition into 30+ day delinquencies - by loan type
dqrates <- read_excel(path = dtapath, sheet = "Page 13 Data", skip = 4, col_names = TRUE) %>% rename(timeper=`...1`) 
dqrates$timeper <- as.yearqtr(sapply(dqrates$timeper, dtfmt))
#fix timeper from before 2011
dqrates[1:32,"timeper"] <- c(as.yearqtr("2003 Q1"), as.yearqtr("2003 Q2"), as.yearqtr("2003 Q3"), as.yearqtr("2003 Q4"),
                             as.yearqtr("2004 Q1"), as.yearqtr("2004 Q2"), as.yearqtr("2004 Q3"), as.yearqtr("2004 Q4"),
                             as.yearqtr("2005 Q1"), as.yearqtr("2005 Q2"), as.yearqtr("2005 Q3"), as.yearqtr("2005 Q4"),
                             as.yearqtr("2006 Q1"), as.yearqtr("2006 Q2"), as.yearqtr("2006 Q3"), as.yearqtr("2006 Q4"),
                             as.yearqtr("2007 Q1"), as.yearqtr("2007 Q2"), as.yearqtr("2007 Q3"), as.yearqtr("2007 Q4"),
                             as.yearqtr("2008 Q1"), as.yearqtr("2008 Q2"), as.yearqtr("2008 Q3"), as.yearqtr("2008 Q4"),
                             as.yearqtr("2009 Q1"), as.yearqtr("2009 Q2"), as.yearqtr("2009 Q3"), as.yearqtr("2009 Q4"),
                             as.yearqtr("2010 Q1"), as.yearqtr("2010 Q2"), as.yearqtr("2010 Q3"), as.yearqtr("2010 Q4"))

#all loan types 30+ day transition delinquencies graph
dqratesmelt <- dqrates %>% select(1:4,6) %>% 
  rename('AUTO' = 'AUTO...2', 'CC' = 'CC...3', 'MORTGAGE' = 'MORTGAGE...4', 'STUDENT LOAN' = 'STUDENT LOAN...6') %>% 
  melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(dqratesmelt, aes(x=timeper, y=value/100, group=variable, color=variable)) +
  geom_line(size=1.5) +
  labs(
    title = "Percent of US Debt Balances Moving from Current to 30+ Days Delinquent, by Loan Type"
    , x = "Year : Quarter"
    , y = "% Balance Transitioning to 30+ Days Delinquent"
    , fill = "Loan Type"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=percent) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5)]) +
  theme(axis.title=element_text(size=10,face="bold"))
# addlogo()

```

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#Transition into 30+ day delinquencies - just credit card
dqrates <- read_excel(path = dtapath, sheet = "Page 13 Data", skip = 4, col_names = TRUE) %>% rename(timeper=`...1`) 
dqrates$timeper <- as.yearqtr(sapply(dqrates$timeper, dtfmt))
#fix timeper from before 2011
dqrates[1:32,"timeper"] <- c(as.yearqtr("2003 Q1"), as.yearqtr("2003 Q2"), as.yearqtr("2003 Q3"), as.yearqtr("2003 Q4"),
                             as.yearqtr("2004 Q1"), as.yearqtr("2004 Q2"), as.yearqtr("2004 Q3"), as.yearqtr("2004 Q4"),
                             as.yearqtr("2005 Q1"), as.yearqtr("2005 Q2"), as.yearqtr("2005 Q3"), as.yearqtr("2005 Q4"),
                             as.yearqtr("2006 Q1"), as.yearqtr("2006 Q2"), as.yearqtr("2006 Q3"), as.yearqtr("2006 Q4"),
                             as.yearqtr("2007 Q1"), as.yearqtr("2007 Q2"), as.yearqtr("2007 Q3"), as.yearqtr("2007 Q4"),
                             as.yearqtr("2008 Q1"), as.yearqtr("2008 Q2"), as.yearqtr("2008 Q3"), as.yearqtr("2008 Q4"),
                             as.yearqtr("2009 Q1"), as.yearqtr("2009 Q2"), as.yearqtr("2009 Q3"), as.yearqtr("2009 Q4"),
                             as.yearqtr("2010 Q1"), as.yearqtr("2010 Q2"), as.yearqtr("2010 Q3"), as.yearqtr("2010 Q4"))

#all loan types 30+ day transition delinquencies graph
dqratesmelt <- dqrates %>% select(1:4,6) %>% 
  rename('AUTO' = 'AUTO...2', 'CC' = 'CC...3', 'MORTGAGE' = 'MORTGAGE...4', 'STUDENT LOAN' = 'STUDENT LOAN...6') %>% 
  melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(dqratesmelt%>%filter(timeper>='2015 Q1', variable=='CC'), aes(x=timeper, y=value/100, group=1)) +
  geom_line(color=wharton_palette[7], size=1.5) +
  # geom_area(fill=wharton_palette[7], alpha=0.5) +
  labs(
    title = "Percent of US Credit Card Balances Moving from Current to 30+ Days Delinquent"
    , x = "Year : Quarter"
    , y = "% Balance Transitioning to 30+ Days Delinquent"
    , fill = "Loan Type"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=percent) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5)]) +
  theme(axis.title=element_text(size=10,face="bold"))
# addlogo()

```

“Macy’s bonds slide amid consumer pullback, surge in credit card delinquencies.”
Why are consumer finances stressed when everybody has a job?


```{r, results = TRUE, echo = FALSE, warning=FALSE}
# Unemployment rate
getSymbols.FRED("UNRATE", env = .GlobalEnv) 

unrate <- as.data.frame(UNRATE)
unrate$dt <- as.Date(row.names(unrate))

# unrate$yrqtr <- as.yearqtr(unrate$dt)

ggplot(unrate%>%filter(dt>='2010-01-01'), aes(x=dt, y=UNRATE/100, group=1)) +
  geom_line(color=wharton_palette[4], size=1.5) +
  # geom_point(data = ueclaims%>%filter(dt >= '2020-04-01'), color=ocrolus_palette[7]) +
  # geom_text_repel(data = ueclaims%>%filter(dt >= '2020-04-01'),
  #                 aes(label=paste0(strftime(dt, '%b %d'), ": ", comma(CCSA))),
  #                 size = 3.5, family='Source Sans Pro') +
  labs(
    title = "National Unemployment Rate in the United States"
    , x = "Year"
    , y = "Unemployment Rate"
    , caption = "Source: Board of Governors of the Federal Reserve System"
  ) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(labels=percent) +
  scale_x_date(date_labels = '%Y', date_breaks = '1 year') +
  # scale_x_yearqtr(format = '%Y-Q%q') +
  wharton_theme 
# addlogo()

```

However, consumers aren’t moving into financial distress at the same level they did when card or auto delinquencies were last occurring at those paces. The number of consumers facing new foreclosures and bankruptcies is still at levels well below the prepandemic period. The New York Fed noted in its report that foreclosures have stayed low even after a national pause on foreclosures that was included in the 2020 Coronavirus Aid, Relief, and Economic Security Act, or Cares Act, ended in 2021. Some forms of foreclosure relief or limitations continued.

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#number of consumers with new foreclosures and bankruptcies
foreclosures <- read_excel(path = dtapath, sheet = "Page 17 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`...1`) %>% select(1:3)
foreclosures$timeper <- as.yearqtr(sapply(foreclosures$timeper, dtfmt))
foreclosuresmelt <- foreclosures %>% melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(foreclosuresmelt, aes(x=timeper, y=value, fill=variable)) +
  geom_col() +
  labs(
    title = "Number of Consumers with New Foreclosures & Bankruptcies"
    , x = "Year : Quarter"
    , y = "Thousands ($)"
    , fill = "Delinquency Status"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=dollar) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5,6,7)]) 
# addlogo()

```

30 year yield

```{r, results = TRUE, echo = FALSE, warning=FALSE}
# 30 year fixed rate mortgage
getSymbols.FRED("MORTGAGE30US", env = .GlobalEnv) 

thirty_year <- as.data.frame(MORTGAGE30US)
thirty_year$dt <- as.Date(row.names(thirty_year))

# thirty_year$yrqtr <- as.yearqtr(thirty_year$dt)

ggplot(thirty_year%>%filter(dt>='2018-01-01'), aes(x=dt, y=MORTGAGE30US/100, group=1)) +
  geom_line(color=wharton_palette[4], size=2) +
  labs(
    title = "30-Year Fixed Rate Mortgage Average in the United States"
    , x = "Year"
    , y = "Rate"
    , caption = "Source: Board of Governors of the Federal Reserve System"
  ) +
  scale_y_continuous(labels=percent) +
  scale_x_date(date_labels = '%Y', date_breaks = '1 year') +
  # scale_x_yearqtr(format = '%Y-Q%q') +
  wharton_theme 
# addlogo()


```

```{r, results = TRUE, echo = FALSE, warning=FALSE}
# 30 year fixed rate mortgage
getSymbols.FRED("MORTGAGE30US", env = .GlobalEnv) 

thirty_year <- as.data.frame(MORTGAGE30US)
thirty_year$dt <- as.Date(row.names(thirty_year))

# thirty_year$yrqtr <- as.yearqtr(thirty_year$dt)

ggplot(thirty_year%>%filter(dt>='2003-01-01'), aes(x=dt, y=MORTGAGE30US/100, group=1)) +
  geom_line(color=wharton_palette[4], size=1.5) +
  labs(
    title = "30-Year Fixed Rate Mortgage Average in the United States"
    , x = "Year"
    , y = "Rate"
    , caption = "Source: Board of Governors of the Federal Reserve System"
  ) +
  scale_y_continuous(labels=percent) +
  scale_x_date(date_labels = '%Y', date_breaks = '3 years') +
  # scale_x_yearqtr(format = '%Y-Q%q') +
  theme(axis.text.x = element_text(angle = 90)) +
  wharton_theme 
# addlogo()


```

Mortgage originations way down

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#new mortgage loans by fico
newmortgageloans <- read_excel(path = dtapath, sheet = "Page 6 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`...1`) %>% select(1:6)
names(newmortgageloans) <- c('timeper', '< 620', '620-659', '660-719', '720-759', '760+')
newmortgageloans$timeper <- as.yearqtr(sapply(newmortgageloans$timeper, dtfmt))
newmortgageloans <- newmortgageloans %>% melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(newmortgageloans, aes(x=timeper, y=value, fill=variable)) +
  geom_col() +
  labs(
    title = "New Mortgage Originations by Credit Score Range"
    , x = "Year : Quarter"
    , y = "New Loan Origination ($ billion)"
    , fill = "Credit Score"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=dollar) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5,6,7)]) 
# addlogo()

```


## Indicators of Recovery

securitizations of fixed coupon bonds

```{r, results = TRUE, echo = FALSE, warning=FALSE}
# Costs of Funds of Small Business Lending (Securitizations with Fixed-Coupon Bonds)
df <- data.frame (issuer_rating  = c("Channel Partners-A", "Channel Partners-BBB",
                                     "Kapitus-A", "Kapitus-AA", "Kapitus-BBB",
                                     # "National Funding-A", "National Funding-BBB-",
                                     "Rapid Finance-A", "Rapid Finance-AA", "Rapid Finance-AAA", "Rapid Finance-BBB",
                                     "Enova Intl-A", "Enova Intl-AAA", "Enova Intl-BBB",
                                     "Channel Partners-A", "Channel Partners-BBB",
                                     "Kapitus-A", "Kapitus-AA", "Kapitus-BBB",
                                     # "National Funding-A", "National Funding-BBB-",
                                     "Rapid Finance-A", "Rapid Finance-AA", "Rapid Finance-AAA", "Rapid Finance-BBB",
                                     "Enova Intl-A", "Enova Intl-AAA", "Enova Intl-BBB"),
                  issue_date = c("12-01-2021", "12-01-2021",
                                 "02-01-2022", "02-01-2022", "02-01-2022",
                                 # "08-21-2019", "08-21-2019",
                                 "07-01-2021", "07-01-2021", "07-01-2021", "07-01-2021",
                                 "04-27-2021", "04-27-2021", "04-27-2021",
                                 "06-27-2023", "06-27-2023",
                                 "12-01-2022", "12-01-2022", "12-01-2022",
                                 # "09-19-2022", "09-19-2022",
                                 "11-01-2022", "11-01-2022", "11-01-2022", "11-01-2022",
                                 "07-20-2023", "07-20-2023", "07-20-2023"),
                  coupon = c(.0358, .0407,
                             .0358, .0338, .0446,
                             # .0417, .0506,
                             .0260, .0230, .0196, .0344,
                             .0228, .0159, .0297,
                             .0898, .1138,
                             .0787, .0690, .1147,
                             # .0759, .1161,
                             .0805, .0757, .0708, .0950,
                             .0825, .0700, .0993)
)

df$issue_date <- mdy(df$issue_date)

# securitizations  with fixed coupon bonds
ggplot(df, aes(x=issue_date, y=coupon, color=issuer_rating, group=issuer_rating)) +
  geom_line() +
  geom_point() +
  scale_x_date(date_labels = '%Y-%m', date_breaks = '3 months') +
  wharton_theme_dark +
  # wharton_theme +
  scale_y_continuous(labels=percent) +
  labs(
    title = "Cost of Funds for Small Business Lending \nSecuritizations with Fixed-Coupon Bonds"
    , x = "Issue Date"
    , y = "Coupon"
    , color = "Issuer & Rating"
    , caption = "Selected Deals, Source: finsight.com"
  ) +
  theme(
    legend.position = 'right', 
    legend.text = element_text(size = 10)
    #legend.key = element_blank(), 
    #legend.key.size = unit(1,"cm")
  ) +
  theme(axis.text = element_text(size = 8))
# addlogo()

```

Tightening standards

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#tightening standards graph
getSymbols.FRED(c('DRTSCLCC', 'STDSAUTO', 'DRTSCIS', 'DRTSCILM', 'DRTSSP', 'SUBLPDRCSC', 'SUBLPDHMSENQ', 'SUBLPDHMSGNQ'), env = .GlobalEnv)

cons <- as.data.frame(DRTSCLCC) %>% rename(pct = DRTSCLCC)
cons$dt <- as.Date(row.names(cons))
cons$desc <- 'Consumer Loans & Credit Cards'

auto <- as.data.frame(STDSAUTO) %>% rename(pct = STDSAUTO)
auto$dt <- as.Date(row.names(auto))
auto$desc <- 'Auto Loans'

commind <- as.data.frame(DRTSCIS) %>% rename(pct = DRTSCIS)
commind$dt <- as.Date(row.names(commind))
commind$desc <- 'Commercial & Industrial Loans to Small Firms'

commind_large <- as.data.frame(DRTSCILM) %>% rename(pct = DRTSCILM)
commind_large$dt <- as.Date(row.names(commind_large))
commind_large$desc <- 'Commercial & Industrial Loans to Large & Middle Market Firms'

subprime_mort <- as.data.frame(DRTSSP) %>% rename(pct = DRTSSP)
subprime_mort$dt <- as.Date(row.names(subprime_mort))
subprime_mort$desc <- 'Subprime Mortgage Loans'

gse_mort <- as.data.frame(SUBLPDHMSENQ) %>% rename(pct = SUBLPDHMSENQ)
gse_mort$dt <- as.Date(row.names(gse_mort))
gse_mort$desc <- 'GSE-Eligible Mortgage Loans'

govt_mort <- as.data.frame(SUBLPDHMSGNQ) %>% rename(pct = SUBLPDHMSGNQ)
govt_mort$dt <- as.Date(row.names(govt_mort))
govt_mort$desc <- 'Government Mortgage Loans'

commre_construction <- as.data.frame(SUBLPDRCSC) %>% rename(pct = SUBLPDRCSC)
commre_construction$dt <- as.Date(row.names(commre_construction))
commre_construction$desc <- 'Commercial Real Estate Loans for Construction & Land Development'

alldta <- rbind(cons, auto, commind)
# alldta <- rbind(commind)

ggplot(alldta%>%filter(dt>='2016-01-01'), aes(x=dt, y=pct, color=desc, group=desc)) +
  geom_line(size=1.5) +
  scale_x_date(date_labels = '%Y-%m', date_breaks = '1 year') +
  # wharton_theme_dark +
  wharton_theme +
  labs(
    title = "Net Percentage of Domestic Banks Tightening Underwriting Standards"
    , x = "Period (Quarterly, EOQ)"
    , y = "Percent, Not Seasonally Adjusted"
    , color = "Loan Type"
    , caption = "https://fred.stlouisfed.org/series/DRTSCLCC\nhttps://fred.stlouisfed.org/series/STDSAUTO\nhttps://fred.stlouisfed.org/series/DRTSCIS"
  ) +
  theme(
    legend.position = 'bottom', 
    legend.text = element_text(size = 8),
    #legend.key = element_blank(), 
    #legend.key.size = unit(1,"cm"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14)
  )

```

Reporting stronger demand

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#reporting stronger demand graph
getSymbols.FRED(c('DEMAUTO', 'DEMCC', 'DRSDCIS', 'DRSDSP', 'DRSDCILM', 'SUBLPDHMDENQ', 'SUBLPDHMDGNQ'), env = .GlobalEnv)

cc_dem <- as.data.frame(DEMCC) %>% rename(pct = DEMCC)
cc_dem$dt <- as.Date(row.names(cc_dem))
cc_dem$desc <- 'Credit Card Loans'

auto_dem <- as.data.frame(DEMAUTO) %>% rename(pct = DEMAUTO)
auto_dem$dt <- as.Date(row.names(auto_dem))
auto_dem$desc <- 'Auto Loans'

comm_dem <- as.data.frame(DRSDCIS) %>% rename(pct = DRSDCIS)
comm_dem$dt <- as.Date(row.names(comm_dem))
comm_dem$desc <- 'Commercial & Industrial Loans to Small Firms'

comm_large <- as.data.frame(DRSDCILM) %>% rename(pct = DRSDCILM)
comm_large$dt <- as.Date(row.names(comm_large))
comm_large$desc <- 'Commercial & Industrial Loans to Large Firms'

subprime_mort_dem <- as.data.frame(DRSDSP) %>% rename(pct = DRSDSP)
subprime_mort_dem$dt <- as.Date(row.names(subprime_mort_dem))
subprime_mort_dem$desc <- 'Subprime Mortgage Loans'

gse_mort_dem <- as.data.frame(SUBLPDHMDENQ) %>% rename(pct = SUBLPDHMDENQ)
gse_mort_dem$dt <- as.Date(row.names(gse_mort_dem))
gse_mort_dem$desc <- 'GSE-Eligible Mortgage Loans'

govt_mort_dem <- as.data.frame(SUBLPDHMDGNQ) %>% rename(pct = SUBLPDHMDGNQ)
govt_mort_dem$dt <- as.Date(row.names(govt_mort_dem))
govt_mort_dem$desc <- 'Government Mortgage Loans'

alldta_dem <- rbind(cc_dem, auto_dem, comm_dem)
# alldta_dem <- rbind(comm_dem)

ggplot(alldta_dem%>%filter(dt>='2016-01-01'), aes(x=dt, y=pct, color=desc, group=desc)) +
  geom_line(size=1.5) +
  scale_x_date(date_labels = '%Y-%m', date_breaks = '1 year') +
  # wharton_theme_dark +
  wharton_theme +
  labs(
    title = "Net Percentage of Domestic Banks Reporting Stronger Demand"
    , x = "Period (Quarterly, EOQ)"
    , y = "Percent, Not Seasonally Adjusted"
    , color = "Loan Type"
    , caption = "https://fred.stlouisfed.org/series/DRTSCLCC\nhttps://fred.stlouisfed.org/series/STDSAUTO\nhttps://fred.stlouisfed.org/series/DRTSCIS"
  ) +
  theme(
    legend.position = 'bottom', 
    legend.text = element_text(size = 8),
    #legend.key = element_blank(), 
    #legend.key.size = unit(1,"cm")
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14)
  )


```


Auto getting stronger

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#new auto loans by fico
newautoloans <- read_excel(path = dtapath, sheet = "Page 8 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`...1`) %>% select(1:6)
names(newautoloans) <- c('timeper', '< 620', '620-659', '660-719', '720-759', '760+')
newautoloans$timeper <- as.yearqtr(sapply(newautoloans$timeper, dtfmt))
newautomelt <- newautoloans %>% melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(newautomelt, aes(x=timeper, y=value, fill=variable)) +
  geom_col() +
  labs(
    title = "New Auto Loan Origination by Credit Score Range"
    , x = "Year : Quarter"
    , y = "New Loan Origination ($ billion)"
    , fill = "Credit Score"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=dollar) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5,6,7)]) 
# addlogo()

```

different than last time

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#90+ day delinquencies - recession-level highs
dqrates <- read_excel(path = dtapath, sheet = "Page 12 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`...1`) 
dqrates$timeper <- as.yearqtr(sapply(dqrates$timeper, dtfmt))
#fix timeper from before 2011
dqrates[1:32,"timeper"] <- c(as.yearqtr("2003 Q1"), as.yearqtr("2003 Q2"), as.yearqtr("2003 Q3"), as.yearqtr("2003 Q4"),
                             as.yearqtr("2004 Q1"), as.yearqtr("2004 Q2"), as.yearqtr("2004 Q3"), as.yearqtr("2004 Q4"),
                             as.yearqtr("2005 Q1"), as.yearqtr("2005 Q2"), as.yearqtr("2005 Q3"), as.yearqtr("2005 Q4"),
                             as.yearqtr("2006 Q1"), as.yearqtr("2006 Q2"), as.yearqtr("2006 Q3"), as.yearqtr("2006 Q4"),
                             as.yearqtr("2007 Q1"), as.yearqtr("2007 Q2"), as.yearqtr("2007 Q3"), as.yearqtr("2007 Q4"),
                             as.yearqtr("2008 Q1"), as.yearqtr("2008 Q2"), as.yearqtr("2008 Q3"), as.yearqtr("2008 Q4"),
                             as.yearqtr("2009 Q1"), as.yearqtr("2009 Q2"), as.yearqtr("2009 Q3"), as.yearqtr("2009 Q4"),
                             as.yearqtr("2010 Q1"), as.yearqtr("2010 Q2"), as.yearqtr("2010 Q3"), as.yearqtr("2010 Q4"))

#just auto delinquencies graph (geom area)
ggplot(dqrates, aes(x=timeper, y=AUTO/100, group=1)) +
  geom_area(fill=wharton_palette[4]) +
  labs(
    title = "Auto Loans - % of Balance 90+ Days Past Due"
    , x = "Year : Quarter"
    , y = "% of Balance 90+ DPD"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=percent) +
  scale_x_yearqtr(format = '%Y-Q%q') +
  wharton_theme 
# addlogo()

#just cc delinquencies graph (geom line)
ggplot(dqrates, aes(x=timeper, y=CC/100, group=1)) +
  geom_line(color=wharton_palette[4], size=2) +
  labs(
    title = "CC Loans - % of Balance 90+ Days Past Due"
    , x = "Year : Quarter"
    , y = "% of Balance 90+ DPD"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=percent) +
  # scale_x_date(date_labels = '%Y', date_breaks = '1 year') +
  scale_x_yearqtr(format = '%Y-Q%q') +
  wharton_theme
# addlogo()

```


different than last time

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#all loan types delinquencies graph
dqratesmelt <- dqrates %>% select(-ALL) %>% melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(dqratesmelt, aes(x=timeper, y=value/100, fill=variable)) +
  geom_col() +
  labs(
    title = "Percent of Balance 90+ Days Delinquent by Loan Type"
    , x = "Year : Quarter"
    , y = "Percent of Balance 90+ Days Delinquent"
    , fill = "Loan Type"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=percent) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5,6,7)]) 
# addlogo()

```

delinquency by type of delinquency status

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#total balance by delinquency status
balbydq <- read_excel(path = dtapath, sheet = "Page 11 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`...1`) %>% select(1:7)
# names(newautoloans) <- c('timeper', '< 620', '620-659', '660-719', '720-759', '760+')
balbydq$timeper <- as.yearqtr(sapply(balbydq$timeper, dtfmt))
balbydqmelt <- balbydq %>% select(-Current) %>% melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(balbydqmelt, aes(x=timeper, y=value/100, fill=variable)) +
  geom_col() +
  labs(
    title = "Total Balance by Delinquency Status"
    , x = "Year : Quarter"
    , y = "Percent Delinquent"
    , fill = "Delinquency Status"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=percent) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5,6,7)]) 
# addlogo()
```


total debt balance by age

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#total debt balance by age
debtbyage <- read_excel(path = dtapath, sheet = "Page 20 Data", skip = 3, col_names = TRUE) %>% rename(timeper=`quarter`) %>% select(1:7)
# names(newautoloans) <- c('timeper', '< 620', '620-659', '660-719', '720-759', '760+')
debtbyage$timeper <- as.yearqtr(sapply(debtbyage$timeper, dtfmt))
#get rid of first four rows
debtbyage <- tail(debtbyage,-4)
debtbyagemelt <- debtbyage %>% melt(id.vars = c('timeper')) %>%
  group_by(timeper) %>%
  mutate(
    pctoftot = value / sum(value)
  ) %>%
  ungroup()

ggplot(debtbyagemelt, aes(x=timeper, y=value, fill=variable)) +
  geom_col() +
  labs(
    title = "Total Debt Balance by Age"
    , x = "Year : Quarter"
    , y = "Debt Balance ($ Trillion)"
    , fill = "Borrower Age"
    , caption = "Source: New York Fed Consumer Credit Panel/Equifax"
  ) +
  scale_y_continuous(labels=dollar) +
  wharton_theme + 
  scale_x_yearqtr(format = '%Y-Q%q') +
  scale_fill_manual(values=wharton_palette[c(1,2,4,5,6,7)]) 
# addlogo()

```

debt type faced by age

```{r, results = TRUE, echo = FALSE, warning=FALSE}
#debt share by product type and age (2022Q4)
# For TWO CATEGORICALS (risk tier & industry) + One CONTINUOUS (loss rates units and dollars)
# TIME PERIOD: STATIC
# Facet (on two categorical dimensions)
debtshare <- read_excel(path = dtapath, sheet = "Page 21 Data", skip = 2, col_names = TRUE) %>% rename(credit_type=`...1`) %>% select(1:7)
debtshare <- head(debtshare, -3)

debtsharemelt <- debtshare %>% melt(id.vars=c('credit_type'))

ggplot(debtsharemelt, aes(x=variable, y=value, fill=credit_type)) +
  geom_col(position = 'dodge') +
  scale_y_continuous(labels=dollar) +
  wharton_theme +
  labs(
    title = "Debt Share by Age - Facet by Credit Type"
    , x = "Age Range"
    , y = "Debt ($ Trillion)"
    , fill = "Credit Type"
  ) + 
  scale_fill_manual(values = wharton_palette[c(1,2,3,4,5,6)])  +
  theme(axis.text = element_text(size = 8), legend.text = element_text(size = 8), axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ credit_type)

```
